---
title: "Methods"
subtitle: "Instrumental variables, spatial autocorrelation, and the identification strategy"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(targets)
library(spdep)
library(here)
source(here("R/functions/mapping.R"))
```

## Research Design

The goal is to estimate the **causal effect** of proximity to a Black Lives Matter protest on 2020 general election turnout at the census block group level in Georgia. Proximity is measured as the straight-line distance (miles) from a block group's centroid to the nearest 2020 BLM protest event.

### Why OLS Fails

Simple OLS regression of turnout on distance is biased because **protest locations are not randomly assigned.** Protests clustered in urban, high-resource, high-turnout areas — any positive correlation between protest proximity and turnout may reflect pre-existing demographic differences rather than a protest effect.

### Identification Strategy: Rainfall as Instrument

We use **rainfall on protest days** as an instrumental variable for protest exposure. The logic:

1. **Relevance:** Rain reduces protest size and attendance. Block groups with more rainfall during the protest period were, on average, farther from large protests.
2. **Exclusion:** Rainfall on protest days (months before Election Day) has no direct path to November turnout — it operates *only* through its effect on protest size and location.

This gives us a valid instrument if we are willing to assume that rainfall patterns in May–September 2020 are unrelated to block-group–level political mobilization independent of the protests themselves.

### Causal Diagram

```{mermaid}
flowchart LR
  R["Rainfall\n(instrument)"] -->|first stage| D["Protest\nexposure"]
  D -->|causal effect of interest| T["2020 voter\nturnout"]
  U["Unobserved\nconfounders"] -.->|bias| D
  U -.->|bias| T
  style R fill:#4a90d9,color:#fff
  style D fill:#e67e22,color:#fff
  style T fill:#27ae60,color:#fff
  style U fill:#bbb,color:#333,stroke-dasharray: 5 5
```

Rainfall is excluded from the turnout equation (dashed arrow absent), satisfying the exclusion restriction.

## Formal Model

### First Stage

$$\log(\text{dist}_i + 1) = \alpha_0 + \alpha_1 \log(\text{rainfall}_i + 1) + \alpha_2 \text{pop\_dens}_i + \alpha_3 \text{pct\_black}_i + \nu_i$$

where $\hat{\text{dist}}_i$ (fitted values) is the instrument. The F-statistic on excluded instruments must exceed 10 for the instrument to be considered strong.

### Second Stage (2SLS)

$$\text{turnout}_i = \beta_0 + \beta_1 \widehat{\text{dist}}_i + \sum_t \gamma_t \text{turnout}_{it} + \varepsilon_i$$

$\beta_1$ is the causal estimate of interest: the effect of being one mile farther from a protest on block group turnout share.

### Reduced Form

$$\text{turnout}_i = \delta_0 + \delta_1 \log(\text{rainfall}_i + 1) + \sum_t \gamma_t \text{turnout}_{it} + \eta_i$$

The ratio $\hat\delta_1 / \hat\alpha_1$ gives the Wald (IV) estimate, which should match the 2SLS coefficient.

## Spatial Autocorrelation

Census block groups that are geographically close to each other tend to have similar turnout rates — a violation of the OLS independence assumption. We quantify this with **Moran's I**, a spatially-weighted correlation coefficient:

$$I = \frac{n}{S_0} \frac{\sum_i \sum_j w_{ij}(x_i - \bar{x})(x_j - \bar{x})}{\sum_i (x_i - \bar{x})^2}$$

where $w_{ij}$ is the spatial weight between units $i$ and $j$ (queen contiguity, row-standardized) and $S_0 = \sum_i \sum_j w_{ij}$.

### Spatial Autocorrelation in Key Variables

```{r morans-table}
knitr::kable(
  tar_read(moran_results),
  col.names = c("Variable", "Moran's I", "p-value"),
  digits    = c(0, 3, 4),
  caption   = "Global Moran's I test for spatial autocorrelation (queen contiguity weights)."
)
```

```{r morans-plot, fig.cap="Moran's I scatterplot: 2020 turnout vs. neighbors' average turnout. Positive slope confirms spatial clustering."}
GA_iv      <- tar_read(GA_iv)
ga_weights <- tar_read(ga_weights)

moran.plot(
  GA_iv$General_2020_11_03,
  ga_weights,
  zero.policy = TRUE,
  xlab = "Share of block group turnout (2020 general)",
  ylab = "Spatially lagged turnout (neighbors' average)"
)
```

### Spatial Model Specification

Lagrange multiplier (LM) tests on the baseline OLS model indicate non-independent error terms arising from both spatial lag and spatial error processes. We therefore estimate a **spatial error model**:

$$\text{turnout}_i = X_i \beta + u_i, \quad u_i = \lambda \sum_j w_{ij} u_j + \varepsilon_i$$

which absorbs spatially structured unobservables into the error term, producing consistent coefficient estimates. For the IV setting, we use the predicted distance from the first stage as the treatment in the spatial error framework — a practical implementation of the spatial 2SLS approach outlined in [Betz, Cook & Hollenbach (2019)](https://www.cambridge.org/core/journals/political-science-research-and-methods/article/abs/spatial-interdependence-and-instrumental-variable-models/D00477D4598A9C11BE9B2F2A736366F8).

## Variable Summary

```{r variable-maps, fig.show="hold", out.width="50%", fig.cap="Spatial distributions of distance from protest (left) and relative rainfall (right). Both show clear geographic clustering."}
tar_read(map_distance)
tar_read(map_rainfall)
```
