---
title: "Results"
subtitle: "OLS, IV, and spatial model estimates for Georgia"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(targets)
library(modelsummary)
library(ivreg)
library(spatialreg)
library(here)
source(here("R/functions/model_tables.R"))

ols_m1       <- tar_read(ols_m1)
ols_m3       <- tar_read(ols_m3)
first_stage  <- tar_read(first_stage)
reduced_form <- tar_read(reduced_form)
iv_2sls      <- tar_read(iv_2sls)
sp_error     <- tar_read(sp_error)
lm_tests     <- tar_read(lm_tests)
```

## Main Results

The table below compares four specifications in progression from naive OLS to the spatially-corrected IV model. Reading across columns reveals how the distance estimate changes as we account for confounding (controls), endogeneity (2SLS), and spatial dependence (spatial error).

```{r main-table, results="asis"}
model_table(
  list(
    "OLS (2018 only)"    = ols_m1,
    "OLS (controls)"     = ols_m3,
    "2SLS"               = iv_2sls,
    "Spatial Error"      = sp_error
  ),
  output = "html",
  title  = "Effect of BLM Protest Proximity on 2020 General Election Turnout",
  notes  = "Unit of observation: census block group, Georgia. Outcome: share of registered voters turning out in the 2020 general election. Distance measured in miles from block group centroid to nearest 2020 BLM protest. Standard errors in parentheses. * p < 0.10, ** p < 0.05, *** p < 0.01."
)
```

## First Stage

The instrument (rainfall) must be strongly correlated with the endogenous variable (distance from protest) for 2SLS to be reliable. The first-stage F-statistic — reported below — should exceed 10; values below this threshold indicate a **weak instrument** problem.

```{r first-stage}
fs_fstat <- summary(first_stage)$fstatistic
cat(sprintf(
  "**First-stage F-statistic: %.2f** (threshold for strong instrument: F > 10)\n\n",
  fs_fstat[1]
))
first_stage_table(first_stage, output = "html")
```

## Instrument Validity: Wu-Hausman Test

The Wu-Hausman test evaluates whether OLS is actually inconsistent — i.e., whether distance is truly endogenous. A small p-value rejects the null that OLS and IV estimates are equivalent, supporting the use of IV.

```{r wu-hausman}
iv_diag <- summary(iv_2sls, diagnostics = TRUE)
wh      <- iv_diag$diagnostics["Wu-Hausman", ]
cat(sprintf(
  "Wu-Hausman test statistic: **%.3f** (p = %.4f)\n\n",
  wh["statistic"], wh["p-value"]
))
```

## Spatial Autocorrelation in Residuals

If OLS residuals are spatially clustered, standard errors are underestimated and inference is unreliable. The maps below compare OLS vs. 2SLS residuals — spatial structure in residuals signals remaining spatial dependence not captured by the model.

```{r residual-maps, fig.show="hold", out.width="50%", fig.cap="Left: OLS residuals. Right: 2SLS residuals. Remaining spatial clustering in OLS residuals motivates the spatial error specification."}
tar_read(map_resid_ols)
tar_read(map_resid_iv)
```

## Lagrange Multiplier Specification Tests

```{r lm-tests, results="markup"}
print(lm_tests)
```

Results from these tests determine whether to use a spatial lag model (missing spatially-lagged DV), a spatial error model (non-independent errors), or a Spatial Durbin model (both). Both lag and error statistics are significant, suggesting a Spatial Durbin specification is appropriate for the final model — a direction for future work.

## Robustness

The following checks help assess the sensitivity of results:

- **Alternative control sets:** Using only 2018 turnout vs. all five prior election years
- **Log vs. level distance:** `log(dist + 1)` vs. `dist` in miles — addresses right-skew
- **County fixed effects:** absorb county-level unobservables (requires sufficient within-county variation in rainfall)

These robustness specifications can be added to `R/04-models.R` following the established pipeline pattern.
