---
title: "Black Lives Matter Protests and Voter Turnout"
thanks: Prepared in preparation for the 2021 Annual Meeting of the American Sociological Association. Please do not circulate without permission.
author:
- Cameron Kimble^[Research and Program Associate, Brennan Center for Justice (cameron.kimble@nyu.edu)]
- Kevin Morris^[Researcher, Brennan Center for Justice (kevin.morris@nyu.edu)]
- Kasey Zapatka^[PhD Student, CUNY Graduate Center (kzapatka@gradcenter.cuny.edu)]
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  bookdown::pdf_document2:
    toc: no
    fig_caption: yes
    latex_engine: xelatex
    keep_tex: true
bibliography: "blm_turnout.bib"
link-citations: yes
fontsize: 12pt
header-includes:
    - \usepackage{rotating}
    - \usepackage{setspace}
abstract: |
  In the summer of 2020, Americans took to the streets in larger numbers than ever before. Following the police murders of George Floyd and Breonna Taylor, an enormous multiracial coalition voiced their dissatisfaction with the state of policing in the United States. These mass protests took place shortly before a presidential election, and the incumbent loudly voiced his disdain for protesters and their political message. This paper explores one aspect of the impact of the Black Lives Matter (BLM) movement. Using a national voter file and data on protest location, we aim to estimate the causal impact of physical proximity to a BLM protest on voter turnout. Our pilot studies using Georgia, North Carolina, and Ohio point to a complicated relationship that demands further investigation.
---
\pagenumbering{gobble}
\pagebreak
```{r setup, echo = F, message = F, warning = F}
knitr::opts_knit$set(root.dir = "..", dev = "cairo_pdf")
library(systemfit)
library(kevostools)
library(AER)
library(sandwich)
library(lmtest)
library(scales)
library(ggpubr)
library(tidyverse)
library(extrafont)
library(kableExtra)
library(data.table)
library(stargazer)
loadfonts()
```
\pagenumbering{arabic}
\begin{center}
Black Lives Matter Protests and Voter Turnout

Cameron Kimble, Kevin Morris, and Kasey Zapatka
\end{center}

Political socialization --- that is, the gradual development of an individual’s political preferences, their perceptions of the political world and the state, and the internalization of their society’s norms and behaviors [@Fillieule2013] --- has received increasing attention in recent years. Research about how individuals develop identities as citizens and engage in political behavior is being studied with increasing sophistication by sociologists and political scientists alike. Understanding when residents re-examine the legitimacy of the state and their relationship to it is of great importance to scholars and activists alike, as is an examination of the political consequences. This paper furthers our understanding of how these processes play out by examining whether exposure to Black Lives Matter (BLM) protests in the summer of 2020 increased voter turnout.
Our paper tests one aspect of the emancipatory power of the BLM movement and measures the effect these protests had on voter turnout. If protest exposure led to increased voter turnout, it could provide evidence BLM issues were salient in the minds of voters and represented an important mobilizing issue that should be considered in future elections. We plan to use spatial regression techniques to test if distance from a BLM protest in June 2020 led to increased voter turnout in November.

### Theory and Literature {-}

Foundational research  on political socialization indicates that that initial *frame alignment* (or a “schemata of interpretation”) between an individual and a social movement organization is a precondition for participation [@Snow1986]. That frame alignment is a continuous process that occurs between individuals and social movements, transforming and reinforcing the individual’s ideological orientations. By their very nature, protest movements offer the opportunity for a frame realignment. By encouraging residents to reexamine the legitimacy of the police and state, the Black Lives Matter protests offered the possibility of just this sort of frame realignment.

There is, however, remarkably little literature on the political effects of mass protest in the United States. One study found that participating in the Freedom Summer project “radicalized”volunteers, who were then significantly more likely to participate in social movements in subsequent years [@McAdam1989]. In a similar study, researchers conclude that involvement with social movements through the 1960s and 1970s caused a liberal shift in political orientation, as evidenced by subsequent political engagement such as voting for Jimmy Carter and participating in subsequent demonstrations  [@Sherkat1997].

Other scholars have documented the process by which cultural grievance is used to mobilize groups of voters in the post-New Deal United States [@Leege2002]. When social movements challenge the status quo, mobilization efforts from the associated political party intensifies, thereby increasing voter turnout [@Winders1999].

### The Mobilizing Potential of Injustice Narratives {-}

Although protests can be a key site of political socialization, the Black Lives Matter movement centers on a social issue that has historically been *demobilizing.* Indeed, a whole body of research has documented the legal estrangement [@Bell2017] that grows at the intersection of structurally unjust policies and a “cultural orientation” that views the criminal legal system as illegitimate and irrelevant [@Kirk2011]. This has led to institutional avoidance of many sorts [see @Brayne2014], especially when it comes to voter participation [@Lerman2014]. These demobilizing effects may extend even to individuals who have *not* been formally convicted, but rather are in close contact with those who have been [@Morris2020; but see @White2019]. The relationship, however, is complicated: @Walker2014 shows that individuals in “proximal contact” with individuals who have been convicted become more likely to participate in non-electoral political activities such as attending a protest or donating money to a campaign.

@Walker2020, however, provides key insight into how the demobilizing effects of contact with the legal system can be a *mobilizing* force. Walker argues that, when individuals locate their experience with the legal system in narratives of (racial) injustice, the contact can spur them to take action. The Black Lives Matter movement has aimed to do just this. BLM activists have called attention to the fact that US police forces have a long history of racially discriminatory practices. The widespread protests may have caused individuals in close contact with police and the legal system to understand their contact with the police and legal systems *not* as personal failings but as a result of racial inequity, which Walker’s recent work indicates can be  a mobilizing event.

### Data and Methods {-}

To understand the relationship between protests and turnout, we leverage two primary datasets. First, we employ geocoded data from the US Crisis Monitor^[See https://acleddata.com/download/22846/.] that records the location of every Black Lives Matter protest during 2020..

Our second data source is the national, geocoded voter file made available by L2 Political. The voter file includes a host of information about every registered voter in the country, including their gender, age, partisan affiliation, and race (L2 imputes the characteristics that are not reported in each state’s public voter file). The registered voter file also includes information on voter turnout for the 2020 and previous elections. However, as of writing,  not all states have yet entered their 2020 voter turnout into the registered voter file.

We intend to measure whether how far a voter lived from a Black Lives Matter protest in the summer of 2020 was correlated with their turnout in November. We measure the distance between each voter and the closest protest using geocoded records and test whether this distance is significantly related to turnout after controlling for other relevant characteristics, *including* past turnout.

To be sure, this cannot prove the causal link between exposure to protest and increased turnout. Voters were not “randomly” exposed to protests; the same factors that primed an area to stage a protest may have also primed them to turn out at higher rates in the 2020 election. To identify the causal effect of exposure to protest on voter turnout we will leverage variation in rainfall (relative to historical rainfall) in the week following the murder of George Floyd as an instrument for protest exposure. Following established literature, we expect that rainfall in early June was correlated with protest formation, but not related (via mechanisms other than protest formation) to turnout in November. For a fuller discussion of how rainfall can be used to instrument political protests, see @Madestam2013.

### Preliminary Results {-}

As mentioned above, individual-level turnout records are not yet widely available, but will be published in the first half of 2021. However, records are available for Georgia, North Carolina, and Ohio, which we use as pilot studies.

Figure \ref{fig:map} reports the location of all protests staged in June, 2020, in our three pilot states.

```{r, figures-side, fig.show="hold", out.width="32%", echo = FALSE, message = F, warning = F, fig.cap = "\\label{fig:map}Protest Sites, June 2020", fig.pos = "H"}
p1 <- readRDS("./temp/ga_map_1.rds")
p2 <- readRDS("./temp/nc_map_1.rds")
p3 <- readRDS("./temp/OH_map_1.rds")
print(p1)
print(p2)
print(p3)
```

Table \ref{tab:reg} reports the results of an ordinary least squares regression, where each observation is a registered voter. The dependent variable *Voted in 2020* measures whether a voter participated in the 2020 general election, and the primary dependent variable *Distance* measures how far the voter lived from the nearest June protest in their state. We also control for other individual and neighborhood-level characteristics, including each voter’s own participation history. Robust standard errors are clustered by county.^[These are currently run on random 10 percent samples, stratified by county, due to computing restraints. Eventually, full models will be run using the complete universe of voters.]

\begin{singlespace}
\input{"../temp/big_reg_formatted.tex"}
\end{singlespace}

In both all three states, our preliminary models indicate that voters who lived closer to June protests actually turned out at *lower* rates, even after controlling for other relevant characteristics including each voter’s own vote history, population density, and county fixed effects. While the linear term is not statistically significant in the first North Carolina model, the relationship is significant in the model where the squared term is included. In both Georgia and North Carolina, turnout was about 1 percentage point lower for voters who lived within 5 miles of the closest protest than for other voters. Turnout was 2.8 percentage points lower for these voters in Ohio.

### What Comes Next {-}

Our preliminary results indicate that the relationship between distance to protests and turnout is more complicated than we initially expected. This is perhaps true for both methodological and theoretical reasons. Methodologically, we have not yet implemented the instrumental variable approach that will allow us to more precisely estimate the causal effect of protest exposure. Further, the clustered distribution of protests evident in Figure \ref{fig:map} demonstrates a need for spatial regression techniques  to fully account for spatial autocorrelation. Theoretically, it may be that protest exposure has disparate effects on different voters, and that a single estimate of exposure masks heterogeneous effects. We plan to explore these aspects in our full paper.

\newpage
### References {-}
